sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2023-03-13 doc/paper-jss/paper-jss.Rmd
Copy editing; mostly ref formatting

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{lscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \usepackage{gensymb}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \newunicodechar{˚}{\degree}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  \widowpenalty10000
  \clubpenalty10000
output: rticles::jss_article
---

@@

\begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R} Archive Network (CRAN) at <https://CRAN.R-project.org/package=sdmTMB>:

@@

Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial longitude and latitude `lon` and `lat`.

@@

We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`.
We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package.
We specify the family as Tweedie to account for positive continuous density values that also contain zeros.
An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument).

@@

\begin{figure}
\centering
\includegraphics[width=5.2in]{../../figs/pcod-fig}
\caption[]{Output from the Pacific cod spatial model example in Queen Charlotte Sound, BC, Canada.
(A) SPDE mesh (lines) combined with the trawl survey observations (points). Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. The circle area corresponds to biomass density of Pacific cod caught on individual trawls.
(B) Spatial random field: these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect. (C) Overall prediction: these estimates represent the combination of all fixed and random effects. Crosses illustrate center of gravity (\texttt{get\_cog()}) from a spatiotemporal version; lighter crosses are more recent.
(D) Conditional effect of depth modelled as a penalized smoother.
These predictions are made omitting the spatial random field.
(E) Standardized area-weighted population index derived from a spatiotemporal version (\texttt{get\_index()}). The line represents the estimate and the ribbon indicates a 95\% confidence interval ($\pm$ 2 SEs).}\label{fig:pcod-fig}
\end{figure}

@@

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a].
We provide several approaches to assist this process:
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An \pkg{sdmTMB} model can be passed to the \pkg{tmbstan} package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or perform posterior predictive checks (see `?extract_mcmc`).
(4) The `residuals()` method by default returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
A version that uses Markov chain Monte Carlo (MCMC) sampling of the random effects to avoid this issue is recommended but slower (`?residuals.sdmTMB`).
Simulation-based residuals [@dharma] are also possible.
(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function facilitates simulating data without starting from a fitted model. 
Models can be fit to these simulated data to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models. 
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE [@miller2019] and \pkg{spBayes} since it is a prominent package that can fit related predictive-process models without the SPDE.
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix \ref{app:inla}.
<!-- and \pkg{VAST} (Appendix 5). -->

@@

We ran a simple speed comparison between \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
\pkg{mgcv} was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->

>>>>

---
documentclass: jss
author:
  - name: Sean C. Anderson
    orcid: 0000-0001-9563-1937
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
    email: \email{sean.anderson@dfo-mpo.gc.ca}
  - name: Eric J. Ward
    orcid: 0000-0002-4359-0296
    affiliation: 'Northwest Fisheries Science Center \AND'
    address: |
      | National Oceanic and Atmospheric Administration, National Marine Fisheries Service, Seattle, WA, USA
  - name: Philina A. English
    orcid: 0000-0003-2992-6782
    affiliation: 'Pacific Biological Station'
    address: |
      | Fisheries and Oceans Canada, Nanaimo, BC, Canada
  - name: Lewis A. K. Barnett
    orcid: 0000-0002-9381-8375
    affiliation: 'Alaska Fisheries Science Center'
    address: |
      | National Marine Fisheries Service, National Oceanic and Atmospheric Administration, Seattle, WA, USA
bibliography: refs.bib
title:
  formatted: "\\pkg{sdmTMB}: An \\proglang{R} Package for Fast, Flexible, and User-Friendly Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # If you use tex in the formatted title, also supply version without
  plain:     "sdmTMB: An R Package for Fast, Flexible, and Accessible Generalized Linear Mixed Effects Models with Spatial and Spatiotemporal Random Fields"
  # For running headers, if needed
  short:     "\\pkg{sdmTMB}: geostatistical SPDE-based GLMMs with \\pkg{TMB}"
abstract: >
  Geostatistical data are common across scientific fields and available in increasingly large datasets. However, appropriate models to analyse these data, such as generalised linear mixed effects models with Gaussian random fields, are computationally intensive and challenging to implement for many users. Here, we introduce the \proglang{R} package \pkg{sdmTMB}, which extends the flexible interface familiar to users of \pkg{lme4}, \pkg{glmmTMB}, or \pkg{mgcv} to include spatial and spatiotemporal latent fields using a predictive-process SPDE-(stochastic partial differential equation) based approach. SPDE matrices are constructed with \proglang{R}-\pkg{INLA} and estimation is conducted via maximum marginal likelihood with Template Model Builder (\pkg{TMB}) or via Bayesian inference with \pkg{tmbstan} and \pkg{rstan}. We describe the model and explore case studies that illustrate \pkg{sdmTMB}'s flexibility in implementing penalised smoothers, non-stationary processes (time-varying and spatially varying coefficients), and anisotropy (directionally dependent spatial correlation). Additional functionality includes hurdle models, break-point effects, and out-of-sample cross validation. Finally, we compare the functionality, speed, and interfaces of related software, demonstrating that sdmTMB can be an order of magnitude faster than \proglang{R}-\pkg{INLA}. We hope \pkg{sdmTMB}'s accessible interface will help open this useful class of models to a wider field of geostatistical analysts.
keywords:
  # at least one keyword must be supplied
  formatted: [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
  plain:     [Gaussian Markov random fields (GMRF), generalized linear mixed effects models (GLMM), INLA, SPDE, species distribution modelling, R package, spatio-temporal, spatial-temporal, Template Model Builder]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{bm}
  \usepackage{pdflscape}
  \usepackage{tablefootnote}
  \usepackage{threeparttable}
  \usepackage{booktabs}
  \usepackage{pifont}
  \usepackage{newunicodechar}
  \usepackage{gensymb}
  \newunicodechar{✓}{\ding{51}}
  \newunicodechar{✗}{\ding{55}}
  \newunicodechar{˚}{\degree}
  \DeclareGraphicsExtensions{.pdf,.png,.jpg}
  \widowpenalty10000
  \clubpenalty10000
output: rticles::jss_article
---

@@

\begin{landscape}
\begin{figure}[p]
\centering
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See the sdmTMB \href{https://pbs-assess.github.io/sdmTMB/articles/}{model description vignette} for a full description of the model.
Note that \code{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \code{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \code{(1|g)} denotes a random intercept by group \code{g}, and \code{$\sim$ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

\pkg{sdmTMB} can be installed from the Comprehensive \proglang{R} Archive Network (CRAN) at <https://CRAN.R-project.org/package=sdmTMB>

@@

Importantly, `dependencies = TRUE` includes the \proglang{R}-INLA package  [@rue2009; @lindgren2011; @lindgren2015], which is not on CRAN but is needed for the `make_mesh()` function below.

Here, we illustrate a spatial model fit to Pacific Cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific Cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial longitude and latitude `lon` and `lat`.

@@

We can then fit a model with spatial random fields (`spatial = "on"`) via the function `sdmTMB()`.
We use a penalized smoother for depth as a main effect via `s()` from the \pkg{mgcv} package.
We specify the family as Tweedie [@tweedie1984] to account for positive continuous density values that also contain zeros.
An alternative would be the `delta_gamma()` family [@aitchison1955] or to model catch weight with an offset for log(area swept) (via the `offset` argument).

@@

\begin{figure}
\centering
\includegraphics[width=5.2in]{../../figs/pcod-fig}
\caption[]{Output from the Pacific Cod spatial model example in Queen Charlotte Sound, BC, Canada.
(A) SPDE mesh (lines) combined with the trawl survey observations (points). Finer meshes will be slower to fit but generally increase the accuracy of the SPDE approximation. The circle area corresponds to biomass density of Pacific Cod caught on individual trawls.
(B) Spatial random field: these values are shown in link (log) space and represent spatially correlated deviations that are not accounted for by the depth effect. (C) Overall prediction: these estimates represent the combination of all fixed and random effects. Crosses illustrate center of gravity (\texttt{get\_cog()}) from a spatiotemporal version; lighter crosses are more recent.
(D) Conditional effect of depth modelled as a penalized smoother.
These predictions are made omitting the spatial random field.
(E) Standardized area-weighted population index derived from a spatiotemporal version (\texttt{get\_index()}). The line represents the estimate and the ribbon indicates a 95\% confidence interval ($\pm$ 2 SEs).}\label{fig:pcod-fig}
\end{figure}

@@

Validation and selection of state-space models is challenging, particularly when using the Laplace approximation [@thygesen2017a].
We provide several approaches to assist this process:
(1) The Akaike Information Criterion [AIC, @akaike1974] can be calculated with `AIC()`, although AIC has well-documented biases with mixed-effects models [@liang2008].
(2) Alternatively, k-fold cross validation with `sdmTMB_cv()` can be used with user-specified or randomly chosen folds for model selection [e.g., via expected log predictive density [ELPD], @vehtari2017] or to evaluate goodness of fit according to user-calculated criteria (e.g., mean squared error, area under the curve).
(3) An \pkg{sdmTMB} model can be passed to the \pkg{tmbstan} package [@monnahan2018] to sample from the joint posterior with Stan [@carpenter2017], evaluate the accuracy of the Laplace approximation, or perform posterior predictive checks (see `?extract_mcmc`).
(4) The `residuals()` method by default returns randomized quantile [@dunn1996] or probability integral transform (PIT) [@smith1985] residuals.
<!-- These residuals, however, are calculated as $\hat{y} - y$ where $\hat{y}$ is the sum of the fixed effect estimates and the random effect values that maximize the likelihood conditional on the estimated fixed effects.  -->
For state-space models, these residuals have known statistical issues with the Laplace approximation [@thygesen2017a] but are quick to calculate.
A version that uses Markov chain Monte Carlo (MCMC) sampling of the random effects to avoid this issue [@rufener2021] is recommended but slower (`?residuals.sdmTMB`).
Simulation-based residuals [@dharma] are also possible.
(5) The `simulate.sdmTMB()` method can simulate from fitted models and the `sdmTMB_simulate()` function facilitates simulating data without starting from a fitted model. 
Models can be fit to these simulated data to ensure identifiability, evaluate bias and precision in parameter estimation, or evaluate the consequences of model misspecification.

@@

There are many \proglang{R} packages capable of fitting geostatistical spatial or spatiotemporal models [e.g., @heaton2019].
\pkg{sdmTMB}, \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{spaMM} [@rousset2014] are the most closely related, as they all provide a user interface to SPDE-based GRF models. 
In our software comparison (Table \ref{tab:functionality}), we also include \pkg{mgcv} as it can be adapted to use the SPDE [@miller2019] and \pkg{spBayes} [@finley2007; @finley2015] since it is a prominent package that can fit related predictive-process models without the SPDE.
\pkg{sdmTMB}, \pkg{VAST}, and \pkg{mgcv} can estimate anisotropic covariance whereas \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spBayes} are currently limited to isotropic covariance.
\pkg{sdmTMB} and \pkg{mgcv} focus on univariate response data, whereas \pkg{VAST}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, \pkg{spaMM}, and \pkg{spBayes} extend to multivariate responses with various limitations.
To our knowledge, \pkg{VAST} is the only package to implement spatial [@thorson2015] and spatial dynamic factor analysis [@thorson2016] and spatial empirical orthogonal function (EOF) regression [@thorson2020].
Of these packages, only \pkg{sdmTMB} and \pkg{inlabru} can currently fit threshold (e.g., hockey-stick) covariate relationships.
\pkg{spaMM} is limited to a spatial random field (i.e., does not fit spatiotemporal fields) and \pkg{spBayes} implements spatiotemporal fields, but only as a random walk.
There is considerable variability in the available observation likelihoods across packages (Table \ref{tab:functionality}).
We provide comparisons of the syntax and the reproducibility of results from models fit using \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}, and \pkg{inlabru} in Appendix \ref{app:inla}.
<!-- and \pkg{VAST} (Appendix 5). -->

@@

We ran a simple speed comparison between \pkg{sdmTMB}, \proglang{R}-\pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \proglang{R}-\pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
\pkg{mgcv} was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @openblas and PARDISO: @pardiso7.2a) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->
